Distinguishing Between Spin-Aligned and Isotropic Binary Black Hole Populations Using Gravitational Wave Observations

### 2017-03-23 paper/AlignedIsotropicGW.tex
Updated bibliography.

\citet{2017CQGra..34cLT01V} also studied the possibility of
distinguishing aligned and isotropic angular distributions of \ac{BBH}
spins, but concluded that several hundred sources would be required to
adequately separate the models.  The ``aligned'' population in that
study permitted spins nearly parallel or anti-parallel to the orbital
axis, i.e.\ $\hat{S}\cdot \hat{L} \simeq \pm 1$, in contrast to our
aligned models.  Anti-parallel spins are strongly disfavoured in
models of isolated binary evolution.  \ilya{This would be a good place
  for some review articles about pop.\ synth.}

@@

Although BH spin-orbit alignment is assumed in continuum flux
measurements \citep{2015PhR...548....1M}, some BH XRBs, including GRO
J1655-40 \ilya{and 4U 1543-475?} appear to indicate that the micro
quasar jet, presumably aligned with the BH spin axis, is misaligned
with the orbit.  Moreover, initial stellar spins in massive binaries
have been observed to be misaligned \ilya{[ref]}, though there are
opportunities for realignment during binary evolution, as described
below.

@@

The spin directions of binary black holes formed dynamically through
interactions in dense stellar environments \ilya{[add refs. if not
  discussed elsewhere]} are expected to be isotropic given the absence
of a preferred direction \citep[e.g.,][]{2016ApJ...832L...2R} and the
persistence of an isotropic distribution through post-Newtonian
evolution \citep{2004PhRvD..70l4020S}.

Chemically homogeneous evolution is a possible BBH formation channel
specific to rapidly spinning massive stars aligned by tidal effects
\ilya{[add refs. if not discussed elsewhere]}.  However, wind-driven
mass loss can spin down even some of these stars to the point that
values of $\chi_\textrm{eff}$ as low as $\sim 0.2$ are possible,
though alignment should be preserved \citep{2016AnA...588A..50M}.

The classical isolated binary evolution channel through the
common-envelope (CE) phase \ilya{[add refs. if not discussed
  elsewhere]} yields the following typical formation channel for
binary black holes: possible mass transfer from the primary to the
secondary; the collapse of the primary into a BH; the formation and
ejection of a CE as a consequence of dynamically unstable mass
transfer from the secondary to the primary; possible further mass
transfer from remaining secondary core to primary; collapse of the
secondary into a BH; a merger induced by GW emission
\citep[e.g.,][]{2016Natur.534..512B,Stevenson:2017}.  Interaction
before the collapse of the primary could realign the binary, but the
primary’s spin in the wide (separation $\gtrsim 10$ AU) binary can
easily be misaligned by even moderate \citep{2015MNRAS.450L..85M}
natal kicks.  Realignment of the primary through subsequent mass
accretion is unlikely given the small amount of accreted mass, except
perhaps during hyper-critical accretion during CE \ilya{[add ref. to
  Linden?]}; realignment through dissipation in the possibly warped
accretion disk \citep{1975ApJ...195L..65B} is possible, but calculations
suggest it is unlikely for stellar-mass binaries \citep{KLOP}.

The secondary should tidally align with the orbit before the onset of
mass transfer
\citep[e.g.,][]{2000ApJ...541..319K,2013PhRvD..87j4028G}; however,
most of its angular momentum would be lost to the envelope, which will
be ejected during the CE phase.  The remnant core may or may not be
efficiently re-aligned, depending on the separation between the
secondary’s core and the BH immediately after the CE is ejected
\citep{2016MNRAS.462..844K}.  The secondary’s natal kick will be small
relative to the orbital velocity of the close post-CE binary, so any
further kick-induced misalignment will be moderate
\citep{2016ApJ...832L...2R}.

The analysis above assumed the conservation of angular momentum during
core collapse. However, we know that this may not always be the case.
For example, the magnitude of the angular momentum cannot possibly be
conserved during the collapse to a BH if the pre-collapse
dimensionless spin exceeds $1$.  The direction of the angular momentum
may also change; for example, the misalignment of the secondary pulsar
in the double pulsar \ilya{J0737} by $\sim 135^{\circ}$ relative to
the orbit in the apparent absence of a kick indicated that, at least
for slowly spinning objects, a spin tilt during a supernova may be
possible \citep{2011ApJ...742...81F}.  The dependence of the magnitude
and direction of the spin on its pre-collapse values depends on the
stellar structure (i.e., the level of angular momentum coupling
between the collapsing and ejected material), the supernova explosion
(if any), and the details of the post-collapse fallback process
\ilya{[Check if Levin and Matzner are done]}.

@@

BBHs can form as the end point of isolated binary evolution \cite{refs}, or dynamically in dense stellar environments \cite{refs}. The spin misalignments of BBHs is a signature of their formation mechanism; binaries formed through isolated binary evolution are expected to have their spins aligned with the orbital angular momentum, whilst those formed dynamically are expected to have an isotropic distribution.

>>>>

\citet{2017CQGra..34cLT01V} also studied the possibility of
distinguishing aligned and isotropic angular distributions of \ac{BBH}
spins, but concluded that several hundred sources would be required to
adequately separate the models.  The ``aligned'' population in that
study permitted spins nearly parallel or anti-parallel to the orbital
axis, i.e.\ $\hat{S}\cdot \hat{L} \simeq \pm 1$, in contrast to our
aligned models.  Anti-parallel spins are strongly disfavoured in
models of isolated binary evolution \citep[e.g.,]{Kalogera:2000}.

@@

Although BH spin-orbit alignment is assumed in continuum flux
measurements \citep{2015PhR...548....1M}, some BH XRBs, including GRO
J1655-40 \citep{Martin:2008}, 4U 1543-47 \citep{MorningstarMiller:2014} and particularly V4641 Sgr \citep{Orosz:2001,Martin:2008b} appear to indicate that the micro
quasar jet, presumably aligned with the BH spin axis, is misaligned
with the orbit.  Moreover, initial stellar spins in massive binaries
have been observed to be misaligned \citep[e.g.,][]{Albrecht:2009}, though there are
opportunities for realignment during binary evolution, as described
below.

@@

The spin directions of binary black holes formed dynamically through
interactions in dense stellar environments \citep{SigurdssonHernquist:1993,PZMcMillan:2000,Rodriguez:2015,Stone:2016} 
are expected to be isotropic given the absence
of a preferred direction \citep[e.g.,][]{2016ApJ...832L...2R} and the
persistence of an isotropic distribution through post-Newtonian
evolution \citep{2004PhRvD..70l4020S}.

Chemically homogeneous evolution is a possible BBH formation channel
specific to rapidly spinning massive stars aligned by tidal effects
\citep{MandeldeMink:2016,Marchant:2016}.  However, wind-driven
mass loss can spin down even some of these stars to the point that
values of $\chi_\textrm{eff}$ as low as $\sim 0.2$ are possible,
though alignment should be preserved \citep{Marchant:2016}.

The classical isolated binary evolution channel \citep{TutukovYungelson:1973,TutukovYungelson:1993,Lipunov:1997,Dominik:2015} through the
common-envelope (CE) phase \citep{Ivanova:2013} 
yields the following typical formation channel for
binary black holes: possible mass transfer from the primary to the
secondary; the collapse of the primary into a BH; the formation and
ejection of a CE as a consequence of dynamically unstable mass
transfer from the secondary to the primary; possible further mass
transfer from remaining secondary core to primary; collapse of the
secondary into a BH; a merger induced by GW emission
\citep[e.g.,][]{2016Natur.534..512B,Stevenson:2017}.  Interaction
before the collapse of the primary could realign the binary, but the
primary'۪s spin in the wide (separation $\gtrsim 10$ AU) binary can
easily be misaligned by even moderate \citep{Mandel:2015kicks}
natal kicks.  Realignment of the primary through subsequent mass
accretion is unlikely given the small amount of accreted mass, except
perhaps during hyper-critical accretion during CE \citep{2005ApJ...632.1035O}, though even this accretion is likely to be small according to recent results \citep{MacLeodRamirezRuiz:2015}; realignment through dissipation in the possibly warped
accretion disk \citep{1975ApJ...195L..65B} is possible, but calculations
suggest it is unlikely for stellar-mass binaries \citep{KLOP}.

The secondary should tidally align with the orbit before the onset of
mass transfer
\citep[e.g.,][]{2000ApJ...541..319K,2013PhRvD..87j4028G}; however,
most of its angular momentum would be lost to the envelope, which will
be ejected during the CE phase.  The remnant core may or may not be
efficiently re-aligned, depending on the separation between the
secondary's core and the BH immediately after the CE is ejected
\citep{2016MNRAS.462..844K}.  The secondary'۪s natal kick will be small
relative to the orbital velocity of the close post-CE binary, so any
further kick-induced misalignment will be moderate
\citep{2016ApJ...832L...2R}.

The analysis above assumed the conservation of angular momentum during
core collapse. However, we know that this may not always be the case.
For example, the magnitude of the angular momentum cannot possibly be
conserved during the collapse to a BH if the pre-collapse
dimensionless spin exceeds $1$.  The direction of the angular momentum
may also change; for example, the misalignment of the secondary pulsar
in the double pulsar J0737-3039 by $\sim 135^{\circ}$ relative to
the orbit in the apparent absence of a kick indicated that, at least
for slowly spinning objects, a spin tilt during a supernova may be
possible \citep{2011ApJ...742...81F}.  The dependence of the magnitude
and direction of the spin on its pre-collapse values depends on the
stellar structure (i.e., the level of angular momentum coupling
between the collapsing and ejected material), the supernova explosion
(if any), and the details of the post-collapse fallback process.

@@

As discussed above, BBHs can form as the end point of isolated binary evolution or dynamically in dense stellar environments. The spin misalignments of BBHs is a signature of their formation mechanism; binaries formed through isolated binary evolution are expected to have their spins aligned with the orbital angular momentum, whilst those formed dynamically are expected to have an isotropic distribution.
